\texttt{Davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-09-30 paper/main.tex
teletype Davos throughout (except Fig 1 caption bold text)

\title{Davos: a Python package ``smuggler'' for constructing
  lightweight reproducible notebooks}
\author{Paxton C. Fitzpatrick}
\author{Jeremy R. Manning\corref{cor}}
\ead{Jeremy.R.Manning@Dartmouth.edu}
\cortext[cor]{Corresponding author}
\address{Department of Psychological and Brain Sciences\\Dartmouth College, Hanover, NH 03755}

\begin{abstract}
  Reproducibility is a core requirement of modern scientific research.
  For computational research, reproducibility means that code should
  produce the same results, even when run on different systems.  A
  standard approach to ensuring reproducibility entails packaging a
  project's dependencies along with its primary code base.  Existing
  solutions vary in how deeply these dependencies are specified,
  ranging from virtual environments, to containers, to virtual
  machines.  Each of these existing solutions requires installing or
  setting up a system for running the desired code, increasing the
  complexity and time cost of both sharing and engaging with reproducible
  science. Here, we propose a lighter-weight solution: the
  Davos package.  When used in combination with a
  notebook-based Python project, Davos provides a mechanism
  for specifying the correct versions of the project's
  dependencies directly within the code that requires them,
  and automatically installing them in an isolated environment
  when the code is run. The Davos package further
  ensures that these packages and specific versions are used every
  time the notebook's code is executed.  This enables researchers to
  share a complete reproducible copy of their code within a single
  Jupyter notebook file.
\end{abstract}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=0.65\textwidth]{figs/shareable_code_2d}
\caption{\small \textbf{Systems for sharing code within the Python
  ecosystem.} The $x$-axis denotes the ``burden'' placed on users to
  install and configure the given system (systems placed further to the right, that fall within the redder shading, impose a higher setup cost on the user).  The $y$-axis denotes the
  degree to which the system guarantees that the code will run
  similarly for different users (systems placed higher up, that fall within the bluer shading, offer stronger guarantees).
  From left to right, bottom to top: plain-text \textbf{Python
  scripts} (\texttt{.py} files) provide the most basic ``system''
  for sharing raw code.  Scripts may reference external packages, but
  those packages must be manually installed on other users' systems.
  Further, any checking needed to verify that the correct versions of
  those packages were installed must also be performed manually.
  \textbf{Jupyter notebooks} (\texttt{.ipynb} files) comprise embedded
  text, executable code, and media (including rendered figures, code
  output, etc.). When the \textbf{Davos package} is imported
  into a Jupyter notebook, the notebook's functionality is extended to
  automatically install any required external packages (at their
  correct versions, when specified) into an isolated directory kept separate
  from the user's existing set of packages. \textbf{Virtual environments}
  similarly allow users to install required packages into an isolated
  directory, and \textbf{shareable Python environments} extend this to include
  a specific version of Python itself. However, both of these systems require
  users to create, populate, and manage environments manually. This
  typically entails distributing a configuration file (e.g., a
  \texttt{requirements.txt}, \texttt{pyproject.toml}~\cite{CannEtal16}, or
  \texttt{environment.yml} file) that specifies all project dependencies
  (including version numbers) alongside the primary code base. Users can then
  install a third-party tool~\cite[e.g.,][]{BickEtal07, Eust19, Anac12} to read
  the file and build the environment. \textbf{Containers} provide a means of
  defining an isolated environment that includes a complete operating
  system (independent of the user's operating system), in addition to
  (optionally) specifying a virtual environment or other
  configurations needed to provide the necessary computing
  environment. Containers are typically defined using
  specification files (e.g., a plain-text \texttt{Dockerfile}~\cite{Merk14}) that instruct the
  virtualization engine regarding how to build the containerized
  environment. \textbf{Virtual machines} provide a complete
  hardware-level simulation of the computing environment.  In addition
  to simulating specific hardware, virtual machines (typically
  specified using binary image files) must also define operating
  system-level properties of the computing environment.  $\Delta_1$
  represents the increased reproducibility Davos provides over standard
  Jupyter notebooks for no greater setup cost. $\Delta_2$ represents
  Davos's lower setup cost compared to standard virtual environments
  despite more stringently ensuring reproducibility.}
\label{fig:code-sharing}
\end{figure}

@@

We designed Davos to occupy a ``sweet spot'' within this space.
Davos is a notebook-installable package that adds functionality to the
default notebook experience. Like standard Jupyter notebooks,
Davos-enhanced notebooks allow researchers to include text, executable
code, and media within a single file. No further setup or installation is
required from the user, beyond what is needed to run a standard Jupyter
notebook. And like virtual environments, Davos provides a convenient
mechanism for fully specifying (and installing, as needed) a complete set of
Python dependencies, including specific package versions, which are contained
and isolated from the rest of the user's system.

@@

The Davos package is named after Davos Seaworth, a smuggler referred
to as ``the Onion Knight'' from the series \textit{A Song of Ice and Fire} by
George R. R. Martin~\cite{Mart98}. The \texttt{smuggle} keyword provided by
Davos is a play on Python's \texttt{import} keyword: whereas importing
can load a package into the Python workspace within the existing rules and
frameworks provided by the Python language, ``smuggling'' provides an
alternative that expands the scope and reach of ``importing.'' Like the
character Davos Seaworth (who became famous for smuggling onions through a
blockade on his homeland), the Davos package uses ``onion comments'' to precisely control how
packages are smuggled into the Python workspace.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/package_structure}
\caption{\small \textbf{Package structure.} The Davos package
  comprises two interdependent subpackages.  The \texttt{davos.core}
  subpackage includes modules for parsing \texttt{smuggle} statements
  and onion comments, installing and validating packages, isolating and managing
  installed packages, and configuring Davos's behavior.  The
  \texttt{davos.implementations} subpackage includes
  environment-specific modifications and features that are needed to
  support the core functionality across different notebook-based
  environments.  Individual modules (i.e., \texttt{.py} files) are represented by lime
  rounded rectangles, and arrows denote dependencies (each arrow
  points to a module that imports objects defined in the module at the
  arrow's source).}
\label{fig:package-structure}
\end{figure}

@@

The Davos package consists of two interdependent subpackages
(see Fig.~\ref{fig:package-structure}). The first,
\texttt{davos.core}, comprises a set of modules that
implement the bulk of the package's core
functionality, including pipelines for installing and validating
packages, custom parsers for the \texttt{smuggle} statement (see
Sec.~\ref{subsec:smuggle}) and onion comment (see
Sec.~\ref{subsec:onion}), a system for isolating dependencies of
different projects (see Sec.~\ref{subsec:projects}), and a runtime
interface for configuring Davos's behavior (see Sec.~\ref{subsec:config}).
However, certain critical aspects of this
functionality require (often substantially) different implementations
depending on properties of the notebook environment in which
Davos is used (e.g., whether the frontend is provided by
Jupyter or Google Colaboratory, or which version of
IPython~\cite{PereGran07} is used by the notebook kernel). To deal
with this, environment-dependent components of core features and behaviors
are isolated and abstracted to ``helper functions'' in the
\texttt{davos.implementations} subpackage. This second subpackage
defines multiple, interchangeable versions of each helper function,
organized into modules by the conditions that trigger their use. At
runtime, Davos detects various features in the notebook
environment and selectively imports a single version of each helper
function into the top-level \texttt{davos.implementations} namespace,
allowing \texttt{davos.core} modules to access the proper
implementations for the current notebook environment in a single,
consistent location. An additional benefit of this design is that it
allows maintainers and users to extend Davos to
support new, updated, or custom notebook variants by adding new
\texttt{davos.implementations} modules that define their own versions
of each helper function, modified from existing implementations as
needed.

@@

Functionally, importing Davos in an IPython notebook enables
an additional Python keyword: ``\texttt{smuggle}'' (see
Sec.~\ref{subsec:implementation} for details on how this works).
The \texttt{smuggle} keyword can be used as a drop-in
replacement for Python's built-in \texttt{import} keyword to load
packages, modules, and other objects into the notebook's namespace.
However, whereas \texttt{import} will fail if the requested package is
not installed locally, \texttt{smuggle} statements can handle missing
packages on the fly.  If a smuggled package does not exist in the
user's Python environment, Davos will download and install it automatically,
expose its contents to Python's \texttt{import} machinery, and load it
into the notebook for immediate use.

Importantly, packages installed by Davos are made available for use in the
notebook without affecting the user's Python environment or existing packages.
By default, \texttt{smuggle} statements will install missing packages (and any
missing dependencies of those packages) into a notebook-specific, virtual
environment-like directory called a ``project'' (see
Sec.~\ref{subsec:projects}). In turn, \texttt{smuggle} statements executed in a
particular notebook will preferentially load packages from that notebook's
project directory whenever they are available, rather than searching for them
in the user's main Python environment. In this way, \texttt{smuggle}
statements can be substituted for \texttt{import} statements to automatically
ensure that all packages needed to run a notebook are installed and available
at runtime each time the notebook is run, without risking interfering with
dependencies of the user's other Python programs, or other Davos-enhanced
notebooks.

@@

For greater control over the behavior of \texttt{smuggle} statements, Davos
defines an additional construct called the ``onion comment.'' An onion comment
is a special type of inline comment that may be placed on a line containing a
\texttt{smuggle} statement to customize how Davos searches for the smuggled
package locally and, if necessary, downloads and installs it. Onion comments
follow a simple format based on the ``type comment'' syntax introduced in PEP
484~\cite{vanREtal14}, and are designed to make managing packages with Davos
intuitive and familiar. To construct an onion comment, users provide the name
of the installer program (e.g., \texttt{pip}) and the same arguments one would
use to manually install the package as desired via the command line:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used
when installing it) may differ from its top-level module name (i.e., the name
used when importing it). In such cases, an onion comment can be used to ensure
that Davos installs the proper package if it cannot be found locally:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
Because onion comments may be constructed to specify any aspect of
the installer program's behavior, they provide a mechanism for precisely controlling
how, where, and when smuggled packages are installed. Critically, if an onion
comment includes a version specifier~\cite{CoghStuf13}, Davos will ensure that
the version of the package loaded into the notebook matches the specific
version requested (or satisfies the given version constraints). If the smuggled
package exists locally, Davos will extract its version information from its
metadata and compare it to the specifier provided. If the two are incompatible
(or no local installation is found), Davos will download, install, and load a
suitable version of the package instead:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to \texttt{smuggle} specific VCS references (e.g.,
Git~\cite{TorvHama05} branches, commits, tags, etc.):
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
Davos processes onion comments internally before forwarding arguments to the
installer program. In addition to preventing shared notebooks from executing
arbitrary code in a user's shell, this enables Davos to adjust its behavior
based on how particular flags will affect the behavior of the installer
program. For example, including \texttt{pip}'s \texttt{--no-input} flag will also
temporarily enable Davos's non-interactive mode (see Sec.~\ref{subsec:config}).
Similarly, if an onion comment contains either \texttt{-I}/\texttt{--ignore-installed},
\texttt{-U}/\texttt{--upgrade}, or \texttt{--force-reinstall}, Davos will
install and load a new copy of the smuggled package without first checking
for it locally:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Since the purpose of an onion comment is to describe how a smuggled package
should be installed (if necessary) so that it can be loaded and used
immediately, options that would normally cause the package not to be installed
(such as \texttt{-h}/\texttt{--help} or \texttt{--dry-run}) are disallowed. Additionally,
when using a Davos ``project'' to isolate smuggled packages (the default behavior;
see Sec.~\ref{subsec:projects}), onion comments may not contain options that
would change the package's installation location (such as
\texttt{-t}/\texttt{--target}, \texttt{--root}, or \texttt{--prefix}). However, if
the user disables project-based isolation and specifies \texttt{--target <dir>},
Davos will ensure that \texttt{<dir>} is included in the module search path (i.e.,
\texttt{sys.path}), prepending it if necessary, so the package can be loaded.

@@

To prevent Davos-enhanced notebooks from having unwanted side effects on the user's environment, any packages installed via \texttt{smuggle} statements are automatically isolated using a custom, virtual environment-like system called ``projects.''
Davos projects are similar to standard Python virtual environments (e.g., created with the standard library's \texttt{venv} module or a third-party tool like \texttt{virtualenv}~\cite{BickEtal07}) but with a few noteworthy differences that make them generally lighter-weight and simpler to use.
Like a standard virtual environment, a Davos project consists of a directory (within a hidden \texttt{.davos} folder in the user's home directory) that houses third-party packages needed for a particular Python project, workflow, or task.
However, unlike standard virtual environments, Davos projects do not need to be manually created, activated, or deactivated, and they function to \textit{extend} the user's existing Python environment rather than replace it.

When Davos is imported into a notebook, a project directory for that notebook is automatically created (if it does not exist already).
When \texttt{smuggle} statements within that notebook are executed, any packages (or specific versions of packages) that are not already available in the user's Python environment are installed into the notebook's project directory (along with any missing dependencies of those packages).
During each \texttt{smuggle} statement's execution, Davos also temporarily prepends the notebook's project directory to the module search path so that these project-installed packages are visible when searching for smuggled packages locally, and prioritized over those in the user's main environment.

Thus, rather than constructing fully separate Python environments from scratch, Davos projects work by supplementing the user's runtime environment with any additional packages (or specific package versions) needed to satisfy the dependencies of their corresponding notebooks.
In some cases, this might include every package smuggled into a notebook (e.g., if the notebook is run inside a freshly created, empty virtual environment).
In other cases, the user's environment may already provide all required packages, and the notebook's project directory will go unused (in which case it will be deleted automatically when the notebook kernel is shut down).
Regardless of the extent to which the existing environment is augmented, Davos's project system ensures that all smuggled packages are installed locally and loaded successfully at runtime, while the contents of the user's Python environment are never altered.

Additionally, because \texttt{smuggle} statements in a given notebook are evaluated every time the notebook is run, this design ensures that the notebook's requirements will remain satisfied even if the user's Python environment changes.
For example, suppose a user has \texttt{NumPy}~\cite{HarrEtal20} v1.24.3 installed in their current Python environment and runs a Davos-enhanced notebook that smuggles \texttt{NumPy} with ``\texttt{numpy==1.24.3}'' specified in an onion comment (see Sec.~\ref{subsec:onion}).
Since the user's existing version of the package satisfies this requirement, Davos will load it into the notebook.
But if the user later upgrades their environment's \texttt{NumPy} version to v1.25.0 (perhaps as a result of installing a different package that depends on it) and subsequently re-runs this notebook, the local version will longer satisfy this requirement, so Davos will install \texttt{NumPy} v1.24.3 into the notebook's project directory and load that version instead.
From then on, any further changes to the user's \texttt{Numpy} installation would have no effect on Davos's behavior in this particular notebook, as a satisfactory version now exists in its project directory.
(If the version specified in the onion comment were changed, Davos would update the version installed in the project directory accordingly.)
For efficiency, Davos projects will generally not duplicate dependencies already satisfied by the user's Python environment.
However, if desired, adding \texttt{pip}'s \texttt{--ignore-installed} flag to an onion comment in the notebook will cause Davos to install the smuggled package into the project directory whether or not it already exists locally.

By default, each Davos-enhanced notebook will create and use its own notebook-specific project named for the absolute path to the notebook file.
However, before smuggling its required packages, a notebook may be set to instead use an arbitrarily named, notebook-agnostic project by assigning any (non-empty) string to \texttt{davos.project} (see Sec.~\ref{subsec:config}).
This provides a convenient way for multiple related notebooks that share a common set of requirements to use the same Davos project, by setting \texttt{davos.project} to the same string in each one.
It is also possible (though typically not recommended) to disable Davos's project system and instead install smuggled packages directly into the user's Python environment by setting \texttt{davos.project} to \texttt{None}.

When accessed (unless its value has been set to \texttt{None}), \texttt{davos.project} will evaluate to a \texttt{Project} object that represents the project used by the current notebook (strings assigned to \texttt{davos.project} are converted to \texttt{Project}s internally). This object supports methods for interacting with the current project, including locating its directory within the file system, listing all installed packages' names and versions, changing the project's name, and deleting its contents.
\texttt{Project} instances can also be created and managed programmatically, and Davos provides additional utilities for viewing and working with all existing projects (see Secs.~\ref{subsec:config} and \ref{subsec:toplevel}).

\subsubsection{Configuring and querying Davos}\label{subsec:config}

After importing Davos into a notebook, the top-level \texttt{davos} module exposes a set of attributes whose values determine various aspects of Davos's behavior.
The majority of these are writeable options that can be modified to customize how, where, and when Davos installs smuggled packages (see Sec.~\ref{sec:illustrative-example} for an illustrative example).
These include:

\begin{itemize}
\item \texttt{.active}: This attribute controls whether support for \texttt{smuggle}
  statements and onion comments is enabled (\texttt{True}) or
  disabled (\texttt{False}).  When Davos is first imported,
  \texttt{davos.active} is set to \texttt{True} (see Sec.~\ref{subsec:implementation} for implementation details and additional information).

>>>>

\title{\texttt{Davos}: a Python package ``smuggler'' for constructing
  lightweight reproducible notebooks}
\author{Paxton C. Fitzpatrick}
\author{Jeremy R. Manning\corref{cor}}
\ead{Jeremy.R.Manning@Dartmouth.edu}
\cortext[cor]{Corresponding author}
\address{Department of Psychological and Brain Sciences\\Dartmouth College, Hanover, NH 03755}

\begin{abstract}
  Reproducibility is a core requirement of modern scientific research.
  For computational research, reproducibility means that code should
  produce the same results, even when run on different systems.  A
  standard approach to ensuring reproducibility entails packaging a
  project's dependencies along with its primary code base.  Existing
  solutions vary in how deeply these dependencies are specified,
  ranging from virtual environments, to containers, to virtual
  machines.  Each of these existing solutions requires installing or
  setting up a system for running the desired code, increasing the
  complexity and time cost of both sharing and engaging with reproducible
  science. Here, we propose a lighter-weight solution: the
  \texttt{Davos} package.  When used in combination with a
  notebook-based Python project, \texttt{Davos} provides a mechanism
  for specifying the correct versions of the project's
  dependencies directly within the code that requires them,
  and automatically installing them in an isolated environment
  when the code is run. The \texttt{Davos} package further
  ensures that these packages and specific versions are used every
  time the notebook's code is executed.  This enables researchers to
  share a complete reproducible copy of their code within a single
  Jupyter notebook file.
\end{abstract}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=0.65\textwidth]{figs/shareable_code_2d}
\caption{\small \textbf{Systems for sharing code within the Python
  ecosystem.} The $x$-axis denotes the ``burden'' placed on users to
  install and configure the given system (systems placed further to the right, that fall within the redder shading, impose a higher setup cost on the user).  The $y$-axis denotes the
  degree to which the system guarantees that the code will run
  similarly for different users (systems placed higher up, that fall within the bluer shading, offer stronger guarantees).
  From left to right, bottom to top: plain-text \textbf{Python
  scripts} (\texttt{.py} files) provide the most basic ``system''
  for sharing raw code.  Scripts may reference external packages, but
  those packages must be manually installed on other users' systems.
  Further, any checking needed to verify that the correct versions of
  those packages were installed must also be performed manually.
  \textbf{Jupyter notebooks} (\texttt{.ipynb} files) comprise embedded
  text, executable code, and media (including rendered figures, code
  output, etc.). When the \textbf{Davos package} is imported
  into a Jupyter notebook, the notebook's functionality is extended to
  automatically install any required external packages (at their
  correct versions, when specified) into an isolated directory kept separate
  from the user's existing set of packages. \textbf{Virtual environments}
  similarly allow users to install required packages into an isolated
  directory, and \textbf{shareable Python environments} extend this to include
  a specific version of Python itself. However, both of these systems require
  users to create, populate, and manage environments manually. This
  typically entails distributing a configuration file (e.g., a
  \texttt{requirements.txt}, \texttt{pyproject.toml}~\cite{CannEtal16}, or
  \texttt{environment.yml} file) that specifies all project dependencies
  (including version numbers) alongside the primary code base. Users can then
  install a third-party tool~\cite[e.g.,][]{BickEtal07, Eust19, Anac12} to read
  the file and build the environment. \textbf{Containers} provide a means of
  defining an isolated environment that includes a complete operating
  system (independent of the user's operating system), in addition to
  (optionally) specifying a virtual environment or other
  configurations needed to provide the necessary computing
  environment. Containers are typically defined using
  specification files (e.g., a plain-text \texttt{Dockerfile}~\cite{Merk14}) that instruct the
  virtualization engine regarding how to build the containerized
  environment. \textbf{Virtual machines} provide a complete
  hardware-level simulation of the computing environment.  In addition
  to simulating specific hardware, virtual machines (typically
  specified using binary image files) must also define operating
  system-level properties of the computing environment.  $\Delta_1$
  represents the increased reproducibility \texttt{Davos} provides over standard
  Jupyter notebooks for no greater setup cost. $\Delta_2$ represents
  \texttt{Davos}'s lower setup cost compared to standard virtual environments
  despite more stringently ensuring reproducibility.}
\label{fig:code-sharing}
\end{figure}

@@

We designed \texttt{Davos} to occupy a ``sweet spot'' within this space.
\texttt{Davos} is a notebook-installable package that adds functionality to the
default notebook experience. Like standard Jupyter notebooks,
\texttt{Davos}-enhanced notebooks allow researchers to include text, executable
code, and media within a single file. No further setup or installation is
required from the user, beyond what is needed to run a standard Jupyter
notebook. And like virtual environments, \texttt{Davos} provides a convenient
mechanism for fully specifying (and installing, as needed) a complete set of
Python dependencies, including specific package versions, which are contained
and isolated from the rest of the user's system.

@@

The \texttt{Davos} package is named after Davos Seaworth, a smuggler referred
to as ``the Onion Knight'' from the series \textit{A Song of Ice and Fire} by
George R. R. Martin~\cite{Mart98}. The \texttt{smuggle} keyword provided by
\texttt{Davos} is a play on Python's \texttt{import} keyword: whereas importing
can load a package into the Python workspace within the existing rules and
frameworks provided by the Python language, ``smuggling'' provides an
alternative that expands the scope and reach of ``importing.'' Like the
character Davos Seaworth (who became famous for smuggling onions through a
blockade on his homeland), the \texttt{Davos} package uses ``onion comments'' to precisely control how
packages are smuggled into the Python workspace.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/package_structure}
\caption{\small \textbf{Package structure.} The \texttt{Davos} package
  comprises two interdependent subpackages.  The \texttt{davos.core}
  subpackage includes modules for parsing \texttt{smuggle} statements
  and onion comments, installing and validating packages, isolating and managing
  installed packages, and configuring \texttt{Davos}'s behavior.  The
  \texttt{davos.implementations} subpackage includes
  environment-specific modifications and features that are needed to
  support the core functionality across different notebook-based
  environments.  Individual modules (i.e., \texttt{.py} files) are represented by lime
  rounded rectangles, and arrows denote dependencies (each arrow
  points to a module that imports objects defined in the module at the
  arrow's source).}
\label{fig:package-structure}
\end{figure}

@@

The \texttt{Davos} package consists of two interdependent subpackages
(see Fig.~\ref{fig:package-structure}). The first,
\texttt{davos.core}, comprises a set of modules that
implement the bulk of the package's core
functionality, including pipelines for installing and validating
packages, custom parsers for the \texttt{smuggle} statement (see
Sec.~\ref{subsec:smuggle}) and onion comment (see
Sec.~\ref{subsec:onion}), a system for isolating dependencies of
different projects (see Sec.~\ref{subsec:projects}), and a runtime
interface for configuring \texttt{Davos}'s behavior (see Sec.~\ref{subsec:config}).
However, certain critical aspects of this
functionality require (often substantially) different implementations
depending on properties of the notebook environment in which
\texttt{Davos} is used (e.g., whether the frontend is provided by
Jupyter or Google Colaboratory, or which version of
IPython~\cite{PereGran07} is used by the notebook kernel). To deal
with this, environment-dependent components of core features and behaviors
are isolated and abstracted to ``helper functions'' in the
\texttt{davos.implementations} subpackage. This second subpackage
defines multiple, interchangeable versions of each helper function,
organized into modules by the conditions that trigger their use. At
runtime, \texttt{Davos} detects various features in the notebook
environment and selectively imports a single version of each helper
function into the top-level \texttt{davos.implementations} namespace,
allowing \texttt{davos.core} modules to access the proper
implementations for the current notebook environment in a single,
consistent location. An additional benefit of this design is that it
allows maintainers and users to extend \texttt{Davos} to
support new, updated, or custom notebook variants by adding new
\texttt{davos.implementations} modules that define their own versions
of each helper function, modified from existing implementations as
needed.

@@

Functionally, importing \texttt{Davos} in an IPython notebook enables
an additional Python keyword: ``\texttt{smuggle}'' (see
Sec.~\ref{subsec:implementation} for details on how this works).
The \texttt{smuggle} keyword can be used as a drop-in
replacement for Python's built-in \texttt{import} keyword to load
packages, modules, and other objects into the notebook's namespace.
However, whereas \texttt{import} will fail if the requested package is
not installed locally, \texttt{smuggle} statements can handle missing
packages on the fly.  If a smuggled package does not exist in the
user's Python environment, \texttt{Davos} will download and install it automatically,
expose its contents to Python's \texttt{import} machinery, and load it
into the notebook for immediate use.

Importantly, packages installed by \texttt{Davos} are made available for use in the
notebook without affecting the user's Python environment or existing packages.
By default, \texttt{smuggle} statements will install missing packages (and any
missing dependencies of those packages) into a notebook-specific, virtual
environment-like directory called a ``project'' (see
Sec.~\ref{subsec:projects}). In turn, \texttt{smuggle} statements executed in a
particular notebook will preferentially load packages from that notebook's
project directory whenever they are available, rather than searching for them
in the user's main Python environment. In this way, \texttt{smuggle}
statements can be substituted for \texttt{import} statements to automatically
ensure that all packages needed to run a notebook are installed and available
at runtime each time the notebook is run, without risking interfering with
dependencies of the user's other Python programs, or other \texttt{Davos}-enhanced
notebooks.

@@

For greater control over the behavior of \texttt{smuggle} statements, \texttt{Davos}
defines an additional construct called the ``onion comment.'' An onion comment
is a special type of inline comment that may be placed on a line containing a
\texttt{smuggle} statement to customize how \texttt{Davos} searches for the smuggled
package locally and, if necessary, downloads and installs it. Onion comments
follow a simple format based on the ``type comment'' syntax introduced in PEP
484~\cite{vanREtal14}, and are designed to make managing packages with \texttt{Davos}
intuitive and familiar. To construct an onion comment, users provide the name
of the installer program (e.g., \texttt{pip}) and the same arguments one would
use to manually install the package as desired via the command line:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used
when installing it) may differ from its top-level module name (i.e., the name
used when importing it). In such cases, an onion comment can be used to ensure
that \texttt{Davos} installs the proper package if it cannot be found locally:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
Because onion comments may be constructed to specify any aspect of
the installer program's behavior, they provide a mechanism for precisely controlling
how, where, and when smuggled packages are installed. Critically, if an onion
comment includes a version specifier~\cite{CoghStuf13}, \texttt{Davos} will ensure that
the version of the package loaded into the notebook matches the specific
version requested (or satisfies the given version constraints). If the smuggled
package exists locally, \texttt{Davos} will extract its version information from its
metadata and compare it to the specifier provided. If the two are incompatible
(or no local installation is found), \texttt{Davos} will download, install, and load a
suitable version of the package instead:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to \texttt{smuggle} specific VCS references (e.g.,
Git~\cite{TorvHama05} branches, commits, tags, etc.):
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
\texttt{Davos} processes onion comments internally before forwarding arguments to the
installer program. In addition to preventing shared notebooks from executing
arbitrary code in a user's shell, this enables \texttt{Davos} to adjust its behavior
based on how particular flags will affect the behavior of the installer
program. For example, including \texttt{pip}'s \texttt{--no-input} flag will also
temporarily enable \texttt{Davos}'s non-interactive mode (see Sec.~\ref{subsec:config}).
Similarly, if an onion comment contains either \texttt{-I}/\texttt{--ignore-installed},
\texttt{-U}/\texttt{--upgrade}, or \texttt{--force-reinstall}, \texttt{Davos} will
install and load a new copy of the smuggled package without first checking
for it locally:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Since the purpose of an onion comment is to describe how a smuggled package
should be installed (if necessary) so that it can be loaded and used
immediately, options that would normally cause the package not to be installed
(such as \texttt{-h}/\texttt{--help} or \texttt{--dry-run}) are disallowed. Additionally,
when using a \texttt{Davos} ``project'' to isolate smuggled packages (the default behavior;
see Sec.~\ref{subsec:projects}), onion comments may not contain options that
would change the package's installation location (such as
\texttt{-t}/\texttt{--target}, \texttt{--root}, or \texttt{--prefix}). However, if
the user disables project-based isolation and specifies \texttt{--target <dir>},
\texttt{Davos} will ensure that \texttt{<dir>} is included in the module search path (i.e.,
\texttt{sys.path}), prepending it if necessary, so the package can be loaded.

@@

To prevent \texttt{Davos}-enhanced notebooks from having unwanted side effects on the user's environment, any packages installed via \texttt{smuggle} statements are automatically isolated using a custom, virtual environment-like system called ``projects.''
\texttt{Davos} projects are similar to standard Python virtual environments (e.g., created with the standard library's \texttt{venv} module or a third-party tool like \texttt{virtualenv}~\cite{BickEtal07}) but with a few noteworthy differences that make them generally lighter-weight and simpler to use.
Like a standard virtual environment, a \texttt{Davos} project consists of a directory (within a hidden \texttt{.davos} folder in the user's home directory) that houses third-party packages needed for a particular Python project, workflow, or task.
However, unlike standard virtual environments, \texttt{Davos} projects do not need to be manually created, activated, or deactivated, and they function to \textit{extend} the user's existing Python environment rather than replace it.

When \texttt{Davos} is imported into a notebook, a project directory for that notebook is automatically created (if it does not exist already).
When \texttt{smuggle} statements within that notebook are executed, any packages (or specific versions of packages) that are not already available in the user's Python environment are installed into the notebook's project directory (along with any missing dependencies of those packages).
During each \texttt{smuggle} statement's execution, \texttt{Davos} also temporarily prepends the notebook's project directory to the module search path so that these project-installed packages are visible when searching for smuggled packages locally, and prioritized over those in the user's main environment.

Thus, rather than constructing fully separate Python environments from scratch, \texttt{Davos} projects work by supplementing the user's runtime environment with any additional packages (or specific package versions) needed to satisfy the dependencies of their corresponding notebooks.
In some cases, this might include every package smuggled into a notebook (e.g., if the notebook is run inside a freshly created, empty virtual environment).
In other cases, the user's environment may already provide all required packages, and the notebook's project directory will go unused (in which case it will be deleted automatically when the notebook kernel is shut down).
Regardless of the extent to which the existing environment is augmented, \texttt{Davos}'s project system ensures that all smuggled packages are installed locally and loaded successfully at runtime, while the contents of the user's Python environment are never altered.

Additionally, because \texttt{smuggle} statements in a given notebook are evaluated every time the notebook is run, this design ensures that the notebook's requirements will remain satisfied even if the user's Python environment changes.
For example, suppose a user has \texttt{NumPy}~\cite{HarrEtal20} v1.24.3 installed in their current Python environment and runs a \texttt{Davos}-enhanced notebook that smuggles \texttt{NumPy} with ``\texttt{numpy==1.24.3}'' specified in an onion comment (see Sec.~\ref{subsec:onion}).
Since the user's existing version of the package satisfies this requirement, \texttt{Davos} will load it into the notebook.
But if the user later upgrades their environment's \texttt{NumPy} version to v1.25.0 (perhaps as a result of installing a different package that depends on it) and subsequently re-runs this notebook, the local version will longer satisfy this requirement, so \texttt{Davos} will install \texttt{NumPy} v1.24.3 into the notebook's project directory and load that version instead.
From then on, any further changes to the user's \texttt{Numpy} installation would have no effect on \texttt{Davos}'s behavior in this particular notebook, as a satisfactory version now exists in its project directory.
(If the version specified in the onion comment were changed, \texttt{Davos} would update the version installed in the project directory accordingly.)
For efficiency, \texttt{Davos} projects will generally not duplicate dependencies already satisfied by the user's Python environment.
However, if desired, adding \texttt{pip}'s \texttt{--ignore-installed} flag to an onion comment in the notebook will cause \texttt{Davos} to install the smuggled package into the project directory whether or not it already exists locally.

By default, each \texttt{Davos}-enhanced notebook will create and use its own notebook-specific project named for the absolute path to the notebook file.
However, before smuggling its required packages, a notebook may be set to instead use an arbitrarily named, notebook-agnostic project by assigning any (non-empty) string to \texttt{davos.project} (see Sec.~\ref{subsec:config}).
This provides a convenient way for multiple related notebooks that share a common set of requirements to use the same \texttt{Davos} project, by setting \texttt{davos.project} to the same string in each one.
It is also possible (though typically not recommended) to disable \texttt{Davos}'s project system and instead install smuggled packages directly into the user's Python environment by setting \texttt{davos.project} to \texttt{None}.

When accessed (unless its value has been set to \texttt{None}), \texttt{davos.project} will evaluate to a \texttt{Project} object that represents the project used by the current notebook (strings assigned to \texttt{davos.project} are converted to \texttt{Project}s internally). This object supports methods for interacting with the current project, including locating its directory within the file system, listing all installed packages' names and versions, changing the project's name, and deleting its contents.
\texttt{Project} instances can also be created and managed programmatically, and \texttt{Davos} provides additional utilities for viewing and working with all existing projects (see Secs.~\ref{subsec:config} and \ref{subsec:toplevel}).

\subsubsection{Configuring and querying \texttt{Davos}}\label{subsec:config}

After importing \texttt{Davos} into a notebook, the top-level \texttt{davos} module exposes a set of attributes whose values determine various aspects of \texttt{Davos}'s behavior.
The majority of these are writeable options that can be modified to customize how, where, and when \texttt{Davos} installs smuggled packages (see Sec.~\ref{sec:illustrative-example} for an illustrative example).
These include:

\begin{itemize}
\item \texttt{.active}: This attribute controls whether support for \texttt{smuggle}
  statements and onion comments is enabled (\texttt{True}) or
  disabled (\texttt{False}).  When \texttt{Davos} is first imported,
  \texttt{davos.active} is set to \texttt{True} (see Sec.~\ref{subsec:implementation} for implementation details and additional information).
